---
title: "QC"
output: html_notebook
---

This notebook is used for explotary analysis and quality control of the data Yatsunenko et al (2012). Processed data are used for running the covariance regression.

Output:
X, covariate matrix
Y, response matrix in internal node scale
N, library size matrix in internal node scale
Phi_est, estimated log ratio in internal node scale
phy_tree, phylogenetic tree
adj, adjacency matrix of the phylogenetic tree

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
```


```{r}
library(dplyr)
library(microbiome)
library(phyloseq)
library(igraph)
library(ape)
source("../R/helper.R")
```


```{r}
load("../data/raw/Yatsunenko_QIIME_processed.rda")
```

We focus our analysis on genus-level data. 
```{r}
study850_genus = tax_glom(study850, taxrank = "Genus", NArm = TRUE)
summarize_phyloseq(study850_genus)
tax_table(study850_genus)[, "Genus"] %>% duplicated() %>% sum() # there are duplicated genus
```

A second round of merge to avoid duplicated genus. No duplicated genus for the data "subset110_genus".
```{r}
otu = as(otu_table(study850_genus), "matrix")
lab = tax_table(study850_genus)[,"Genus"]

# Aggregate counts by identical labels
otu_agg = rowsum(otu, group = lab, reorder = TRUE, na.rm = TRUE)

# Build a minimal taxonomy table at Genus level
genus_levels = rownames(otu_agg)
tax_agg = cbind(Genus = genus_levels)
rownames(tax_agg) = genus_levels
tax_agg = as.matrix(tax_agg)

# Recreate a clean genus-level phyloseq (tree is dropped here; reattach if you wish)
study850_genus_clean = phyloseq(otu_table(otu_agg, taxa_are_rows = TRUE),
                       tax_table(tax_agg), sample_data(study850_genus))
```

Since the phylogenetic tree is only available for the subset of 110 people. We only keep genus present in the phylogenetic tree.
```{r}
common_genus =intersect(as.character(tax_table(study850_genus_clean)[, "Genus"]), as.character(tax_table(subset110_genus)[, "Genus"]))
study850_common_genus = prune_taxa(tax_table(study850_genus_clean)[, "Genus"]%in% common_genus, study850_genus_clean)
taxa_names(study850_common_genus) <- as.character(tax_table(study850_common_genus)[, "Genus"])
subset110_common_genus = prune_taxa(tax_table(subset110_genus)[, "Genus"]%in% common_genus, subset110_genus)
taxa_names(subset110_common_genus) <- as.character(tax_table(subset110_common_genus)[, "Genus"])
```



```{r}
study850_common_genus_meta = study850_common_genus@sam_data
study850_common_genus_taxonomy = study850_common_genus@tax_table
study850_common_genus_otu = study850_common_genus@otu_table %>% t()
dim(study850_common_genus_otu)
```

### Explotary analysis
Proportion of zeros:
```{r}
apply(study850_common_genus_otu, 2, function(x){mean(x==0)}) %>% summary()
```

Total number of reads
```{r}
colSums(study850_common_genus_otu) %>% hist(breaks = 50)
```

Check the mean abundance
```{r}
colMeans(study850_common_genus_otu) %>% hist(breaks = 50)
```

### Pre-processing
Filter OTU based on non-zero proportion and total counts
```{r}
study850_common_genus_QC = prune_taxa((apply(study850_common_genus_otu, 2, function(x){mean(x==0)}) < 0.5 & colSums(study850_common_genus_otu) > 100),study850_common_genus)
dim(otu_table(study850_common_genus_QC))
```
530 samples and 90 genus.

### Phylogenetic tree
Obtain the generated binary tree corresponding to the filtered genus, classify internal and leaf nodes, and calculate the distance between nodes on the tree.
```{r}
phy_tree = keep.tip(phy_tree(subset110_common_genus), taxa_names(study850_common_genus_QC))
nodes <- track <- c(phy_tree$edge[1,1])
internal = c(1)
i = 1
while(length(track) > 0){
  node = track[1]
  childern = phy_tree$edge[phy_tree$edge[,1] == node,2]
  if(length(childern) > 0){
    nodes = c(nodes, childern)
    track = c(track, childern)
    internal[i] = 1
    internal = c(internal, numeric(length(childern)))
  }
  track = track[-1]
  i = i + 1
}
# nodes: order of the nodes of breath-first search
# internal: whether each node in nodes is an internal node

phy_tree_igraph = graph_from_edgelist(matrix(c(phy_tree$edge[,1],
                                        phy_tree$edge[,2]),
                                      ncol = 2), directed = TRUE)
adj = as.matrix(as_adjacency_matrix(phy_tree_igraph))
adj = adj[c(nodes[internal == 1], nodes[internal == 0]),
          c(nodes[internal == 1], nodes[internal == 0])]
rownames(adj) <- colnames(adj) <- as.character(c(nodes[internal == 1], nodes[internal == 0]))
adj_sym = adj + t(adj)
dist_matrix = distances(graph_from_adjacency_matrix(adj_sym, mode = "undirected", diag = FALSE))
```

### Covariate preprocessing
```{r}
X <- as(sample_data(study850_common_genus_QC), "data.frame") %>%
  mutate(
    # sex: female = 0, male = 1
    sex = case_when(as.character(SEX) == "male" ~ 1L, 
                    as.character(SEX) == "female" ~ 0L, 
                    as.character(SEX) == "unknown" ~ NA),
    # age: (> 3) = 0, (<= 3) = 1
    age = case_when(!is.na(as.numeric(as.character(AGE))) & as.numeric(as.character(AGE)) <= 3 ~ 0L,
                    !is.na(as.numeric(as.character(AGE))) & as.numeric(as.numeric(as.character(AGE))) > 3 ~ 1L,
                    is.na(as.numeric(as.character(AGE))) ~ NA),
    # country dummies: baseline = Malawi
    usa  = if_else(as.character(COUNTRY) == "GAZ:United States of America", 1L, 0L),
    vene    = if_else(as.character(COUNTRY) == "GAZ:Venezuela",   1L, 0L) 
  ) %>%
  select(sex, age, usa, vene) %>%
  mutate(intercept = 1)

table(X$sex)
table(X$age)
table(X$usa)
table(X$vene)
```

Change the baseline group to USA (skipped).
```{r, eval=FALSE}
X <- as(sample_data(study850_common_genus_QC), "data.frame") %>%
  mutate(
    # sex: female = 0, male = 1
    sex = case_when(as.character(SEX) == "male" ~ 1L, 
                    as.character(SEX) == "female" ~ 0L, 
                    as.character(SEX) == "unknown" ~ NA),
    # age: (> 3) = 0, (<= 3) = 1
    age = case_when(!is.na(as.numeric(as.character(AGE))) & as.numeric(as.character(AGE)) <= 3 ~ 0L,
                    !is.na(as.numeric(as.character(AGE))) & as.numeric(as.numeric(as.character(AGE))) > 3 ~ 1L,
                    is.na(as.numeric(as.character(AGE))) ~ NA),
    # country dummies: baseline = USA
    malawi  = if_else(as.character(COUNTRY) == "GAZ:Malawi", 1L, 0L),
    vene    = if_else(as.character(COUNTRY) == "GAZ:Venezuela",   1L, 0L) 
  ) %>%
  select(sex, age, malawi, vene) %>%
  mutate(intercept = 1)
```


```{r}
table(X$sex)
table(X$age)
table(X$malawi)
table(X$vene)
```

### Convert the data to internal node scale
```{r}
n = ncol(otu_table(study850_common_genus_QC))
p = nrow(otu_table(study850_common_genus_QC))-1
Y <- N <- matrix(NA, nrow = n, ncol = p)
for( i in 1:n ){
  result = CountLeaf2Int(adj = adj, Y = t(otu_table(study850_common_genus_QC))[i,])
  Y[i,] = result$Y
  N[i,] = result$N
}
Phi_est = matrix(NA, nrow=n, ncol=p)
for (i in 1:n){
  for (j in 1:p){
    if (N[i,j]==0){
      Phi_est[i,j]=0
    }
    else{
      tmp=Y[i,j]/N[i,j]
      if (tmp==0){
        Phi_est[i,j]=-5
      }
      else{
        if (tmp==1){
          Phi_est[i,j]=5
        }
        else{
          Phi_est[i,j]=stats::qlogis(tmp)
        }
      }
    }
  }
}
```

### Save processed matrix for method running
```{r}
na_id = rowSums(is.na(X)) > 0
X = X[!na_id,] %>% as.matrix()
Y = Y[!na_id,]
N = N[!na_id,]
Phi_est = Phi_est[!na_id,]
# save(X, Y, N, Phi_est, phy_tree, adj, file = "data/processed/processed_obj")
```

### Normalized leaf-level data (CLR)
```{r}
otu_QC = t(otu_table(study850_common_genus_QC))
otu_QC_pseudo = otu_QC
otu_QC_pseudo[otu_QC_pseudo == 0] = 1
otu_CLR = log(otu_QC_pseudo/apply(otu_QC_pseudo, 1, sum))
otu_CLR = otu_CLR - apply(otu_CLR, 1, mean)
```

### Demo
1. Distribution of feature-wise variance (covariance) with covariates
2. Both leaf and internal node levels

```{r}
library(dplyr)
library(ggplot2)
library(ggpubr)

sex_col  <- "sex"
age_col  <- "age"
usa_col  <- "usa"
vene_col <- "vene"
# normalized_data = Phi_est
normalized_data = otu_CLR

X_df <- as.data.frame(X)

meta <- X_df %>%
  transmute(
    sex = factor(.data[[sex_col]],
                 levels = c(0, 1),
                 labels = c("female", "male")),
    age = factor(.data[[age_col]],
                 levels = c(0, 1),
                 labels = c("<= 3", "> 3")),
    country = case_when(
      .data[[usa_col]] == 1 ~ "USA",
      .data[[vene_col]] == 1 ~ "Venezuela",
      TRUE ~ "Malawi"
    ) %>% factor(levels = c("USA", "Venezuela", "Malawi"))
  )

get_feature_variance_by_group <- function(Phi, group_vec, covariate_name) {
  group_levels <- levels(group_vec)
  
  out <- lapply(group_levels, function(g) {
    idx <- which(group_vec == g)
    v <- apply(Phi[idx, , drop = FALSE], 2, var, na.rm = TRUE)
    data.frame(
      covariate = covariate_name,
      group = g,
      feature = seq_along(v),
      variance = as.numeric(v),
      stringsAsFactors = FALSE
    )
  })
  
  bind_rows(out)
}

df_sex_var     <- get_feature_variance_by_group(normalized_data, meta$sex, "Sex")
df_age_var     <- get_feature_variance_by_group(normalized_data, meta$age, "Age")
df_country_var <- get_feature_variance_by_group(normalized_data, meta$country, "Country")

sex_colors <- c("female" = "#1b9e77", "male" = "#d95f02")
age_colors <- c("<= 3" = "#7570b3", "> 3" = "#e7298a")
cty_colors <- c("USA" = "#66a61e", "Venezuela" = "#e6ab02", "Malawi" = "#a6761d")

p_sex_var <- ggplot(df_sex_var, aes(x = group, y = variance, fill = group)) +
  geom_violin(trim = TRUE, alpha = 0.85, color = "gray25") +
  scale_fill_manual(values = sex_colors, name = "Sex") +
  labs(title = "Sex", x = NULL, y = "Feature-wise variance") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

p_age_var <- ggplot(df_age_var, aes(x = group, y = variance, fill = group)) +
  geom_violin(trim = TRUE, alpha = 0.85, color = "gray25") +
  scale_fill_manual(values = age_colors, name = "Age group") +
  labs(title = "Age", x = NULL, y = "Feature-wise variance") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

p_country_var <- ggplot(df_country_var, aes(x = group, y = variance, fill = group)) +
  geom_violin(trim = TRUE, alpha = 0.85, color = "gray25") +
  scale_fill_manual(values = cty_colors, name = "Country") +
  labs(title = "Country", x = NULL, y = "Feature-wise variance") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

combined_plot_var <- ggarrange(
  p_sex_var, p_age_var, p_country_var,
  ncol = 3, nrow = 1,
  align = "hv", 
  legend = "bottom"
)

print(combined_plot_var)
ggsave(
  filename = "../results/figures/Yatsunenko/leaf_feature_variance_violin.pdf",
  plot = combined_plot_var,
  device = "pdf",
  width = 11,      # inches
  height = 5,    # inches
  units = "in"
)
```

```{r}
get_feature_corr_by_group <- function(Phi, group_vec, covariate_name) {
  group_levels <- levels(group_vec)

  out <- lapply(group_levels, function(g) {
    idx <- which(group_vec == g)
    Xg <- Phi[idx, , drop = FALSE]
    s = c(cor(Xg)[upper.tri(cor(Xg))])

    data.frame(
      covariate = covariate_name,
      group = g,
      corr_stat = as.numeric(s),
      stringsAsFactors = FALSE
    )
  })

 bind_rows(out)
}

df_sex_corr <- get_feature_corr_by_group(normalized_data, meta$sex, "Sex")
df_age_corr <- get_feature_corr_by_group(normalized_data, meta$age, "Age")
df_country_corr <- get_feature_corr_by_group(normalized_data, meta$country, "Country")

p_sex_corr <- ggplot(df_sex_corr, aes(x = group, y = corr_stat, fill = group)) +
  geom_violin(trim = TRUE, alpha = 0.85, color = "gray25") +
  scale_fill_manual(values = sex_colors, name = "Sex") +
  labs(title = "Sex", x = NULL, y = "Pairwise correlation") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

p_age_corr <- ggplot(df_age_corr, aes(x = group, y = corr_stat, fill = group)) +
  geom_violin(trim = TRUE, alpha = 0.85, color = "gray25") +
  scale_fill_manual(values = age_colors, name = "Age group") +
  labs(title = "Age", x = NULL, y = "Pairwise correlation") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

p_country_corr <- ggplot(df_country_corr, aes(x = group, y = corr_stat, fill = group)) +
  geom_violin(trim = TRUE, alpha = 0.85, color = "gray25") +
  scale_fill_manual(values = cty_colors, name = "Country") +
  labs(title = "Country", x = NULL, y = "Pairwise correlation") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

combined_plot_corr <- ggarrange(
  p_sex_corr, p_age_corr, p_country_corr,
  ncol = 3, nrow = 1,
  align = "hv", 
  legend = "bottom"
)

print(combined_plot_corr)
ggsave(
  filename = "../results/figures/Yatsunenko/leaf_feature_correlation_violin.pdf",
  plot = combined_plot_corr,
  device = "pdf",
  width = 11,      # inches
  height = 5,    # inches
  units = "in"
)
```